# Exploratory analysis of game data from metacritic.com
```{r setoptions, echo=FALSE}
opts_chunk$set(results='asis')
opts_chunk$set(warning = FALSE)
opts_chunk$set(message = FALSE)
opts_chunk$set(echo=TRUE)
```

### Loading the data
```{r loaddata}
library(lubridate)
library(modeest)
library(reshape2)
library(ggplot2)
games <- read.csv('../data/metacritic-20141019-152743.csv', na.strings="NA",
                  colClasses=c(release_date="Date"))
```


### Basic properties of a dataset
```{r basicproperties, results='markup'}
str(games)
summary(games)
hist(games$release_date, breaks = 'years')
```

## Exploratory analysis
### Who gives better score on average: users or critics?
```{r scoresdens}
d1 <- density(games$metascore) 
d2 <- density(games$user_score, na.rm=T)
plot(range(d1$x, d2$x), range(d1$y, d2$y), type = "n", xlab = "Score", ylab="Density")
lines(d1, col = "red")
lines(d2, col = "blue")
legend(x= 10, y=0.01, legend = c("metascore", "user_score"), col=c("red", "blue"), lwd=2, lty=1)
```

### How do mean and median scores vary across years?
```{r meanmedscorebydate}
mean_scores_by_date <- aggregate(cbind(metascore, user_score) ~ 
                                 format(release_date, "%Y"), data = games, mean) 
names(mean_scores_by_date)[1] <- "year"
mdf <- melt(mean_scores_by_date, value.name = "mean_score")
ggplot(data=mdf, aes(x=year, y=mean_score, group = variable, fill = variable)) + 
geom_bar(stat = 'identity', position = 'dodge') + coord_cartesian(ylim = c(60, 100)) + 
theme(axis.text.x = element_text(angle = 45, hjust = 1))

sum_score_count_by_date <- aggregate(cbind(critics_reviews_count, user_reviews_count) ~
                                 format(release_date, "%Y"), data = games, sum) 
names(sum_score_count_by_date)[1] <- "year"
mdf <- melt(sum_score_count_by_date, value.name = "sum_score_count")
ggplot(data=mdf, aes(x=year, y=sum_score_count, group = variable, fill = variable)) + 
geom_bar(stat = 'identity', position = 'dodge') + 
theme(axis.text.x = element_text(angle = 45, hjust = 1))

median_score_count_by_date <- aggregate(cbind(critics_reviews_count, user_reviews_count) ~
                                 format(release_date, "%Y"), data = games, median) 
names(median_score_count_by_date)[1] <- "year"
mdf <- melt(median_score_count_by_date, value.name = "median_score_count")
ggplot(data=mdf, aes(x=year, y=median_score_count, group = variable, fill = variable)) + 
geom_bar(stat = 'identity', position = 'dodge') + 
theme(axis.text.x = element_text(angle = 45, hjust = 1))

median_scores_by_date <- aggregate(cbind(metascore, user_score) ~ 
                                 format(release_date, "%Y"), data = games, median) 
names(median_scores_by_date)[1] <- "year"
mdf <- melt(median_scores_by_date, value.name = "median_score")
ggplot(data=mdf, aes(x=year, y=median_score, group = variable, fill = variable)) + 
geom_bar(stat = 'identity', position = 'dodge') + coord_cartesian(ylim = c(65, 100)) + 
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### How do median scores vary across genres?
```{r medscoresbygenre}
median_scores_by_genre <- aggregate(cbind(metascore, user_score) ~ 
                                 genre, data = games, median) 
names(median_scores_by_genre)[1] <- "genre"
mdf <- melt(median_scores_by_genre, value.name = "median_score")
ggplot(data=mdf, aes(x=sapply(strsplit(as.character(genre), " "), "[[", 2), 
                     y=median_score, group = variable, colour = variable)) + 
geom_line(stat = 'identity', position = 'dodge') + 
theme(axis.text.x = element_text(angle = 45, hjust = 1))

t <- transform(median_scores_by_genre, scorediff = abs(metascore-user_score), metascore=NULL, user_score=NULL)
max(t$scorediff)
tstval <- t$genre[which.max(t$scorediff)]
```
tstval = `r unlist(strsplit(as.character(tstval), " "))[2]`
### What is the percentage of games in which critics rated above users, below users or the same as users?
```{r scorediffperc}
underrated <- sum(games$metascore < games$user_score, na.rm = TRUE)
overrated <- sum(games$metascore > games$user_score, na.rm = TRUE)
equal <- sum(games$metascore == games$user_score, na.rm = TRUE)
na <- sum(is.na(games$metascore > games$user_score))

underratedperc <- underrated/(underrated+equal+overrated+na)
overratedperc <- overrated/(underrated+equal+overrated+na)
equalperc <- equal/(underrated+equal+overrated+na)
naperc <- na/(underrated+equal+overrated+na)
```
**Underrated** by critics: **`r underratedperc`**<br> 
**Overrated** by critics: **`r overratedperc`** <br> 
Critics score and user score are **equal**: **`r equalperc`**<br> 
Missing values: **`r naperc`**
### Which publishers are overrated/underrated?
```{r scorediffbytoppubs}
toppubs <- aggregate(user_reviews_count ~ publisher, data=games, 
                     quantile, probs=0.75)
top30pubs <- head(toppubs[order(toppubs$user_reviews_count, decreasing=TRUE),], 
                  n=30)
median_scores_by_publisher <- aggregate(cbind(metascore, user_score) ~ 
                                 publisher, data = games, median) 
names(median_scores_by_publisher)[1] <- "publisher"
median_diff_top_pubs <- median_scores_by_publisher[median_scores_by_publisher$publisher %in% top30pubs$publisher,]
t <- transform(median_diff_top_pubs, scorediff = metascore-user_score, metascore=NULL, user_score=NULL)
t <- transform(t, publisher=reorder(publisher, -scorediff))#t[order(t$scorediff, decreasing=T),]
ggplot(data=t, aes(x=publisher,y=scorediff)) + geom_bar(stat="identity") + 
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r misc}
with(games, plot(year(release_date), user_score))
abline(with(games, lm(user_score ~ year(release_date))))
```
